---
title: "R Exploratory Data Analysis"
output: html_notebook
---


#### Reference
[Link](https://r4ds.had.co.nz/exploratory-data-analysis.html)


#### Variation

* Categorical variable - use bar chart
* Continuous variable - use histogram

```{r categorical}
#categorical variable
ggplot(diamonds, aes(cut))+
  geom_bar()

diamonds %>% 
  count(cut)
```



```{r continuous}
#continuous variable
ggplot(diamonds, aes(carat))+
  geom_histogram(binwidth = 0.5)

diamonds %>% 
  count(cut_width(carat, 0.5))
```


If you wish to overlay multiple histograms in the same plot, better to use `geom_freqpoly()` instead of `geom_histogram()`. Because it use lines instead of bar so makes it easier to compare.

#### Missing Values

Replacing unusual values with missing value by using `mutate()` & `ifelse()`.
```{r ifelse}
diamond2 <- diamonds %>% 
  mutate(y = ifelse(y<3|y>=20, NA, y))
```

ggplot2 never silence the missing values but show the warning that they have been removed from plotting. 
In histogram, missing values are removed when the number of observations in each bin are calculated.In `geom_bar()`, NA is treated as another category.


#### Covariation

* 1) A categorical and a continuous variable

```{r geom_freqpoly}
#common way to plot category vs. continuous variable
ggplot(diamonds, aes(price))+
  geom_freqpoly(aes(color=cut), binwidth=500)
```
 
 
However, the above result is hard to see the difference as they differ too much for each cut categories. Also, y axis is represented by count, so if the categories vary much in count, it is hard to see.
 
Can solve it by density method or `geom_boxplot`.

By density, y-axis represent density, which is the count standardized so that the area under each frequency polygon is one.
```{r density}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

`geom_boxplot` is a popular way in this scenario. 

```{r geom_boxplot}
ggplot(diamonds, aes(cut, price))+
  geom_boxplot()
```


Use `reorder()` to order the categories on the plot.

```{r reorder}
#reorder class based on the median value of hwy
ggplot(mpg, aes(x = reorder(class, hwy, FUN = median), y = hwy))+
  geom_boxplot()
```


* 2) Two categorical variables

By `geom_count()`

```{r geom_count}
ggplot(diamonds, aes(cut, color))+
  geom_count()
```


Another way is compute the count with dplyr and visualize with `geom_tile`.
It’s usually better to use the categorical variable with a larger number of categories or the longer labels on the y axis.

```{r geom_tile}
diamonds %>% 
  count(cut, color) %>% 
  ggplot(aes(color, cut))+
  geom_tile(aes(fill = n))
```


* 3) Two continuous variables

By `geom_point`, but it fails with large dataset.
Can use `geom_hex` and `geom_bin2d`.

```{r geom_bin2d}
ggplot(diamonds, aes(carat, price))+
  geom_bin2d()
```

```{r geom_hex}
install.packages("hexbin")
ggplot(diamonds, aes(carat, price))+
  geom_hex()
```

or can consider to bin one continuous variable into category and then plot against another continuous variable, just like (1 cat & 1 cont) plot.

  * `cut_width`
  * `cut_number`

```{r cut_width}
#cut_width(x, width), divides x into bins of width
ggplot(diamonds, aes(carat, price))+
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

```{r cut_number}
ggplot(diamonds, aes(carat, price))+
  geom_boxplot(aes(group = cut_number(carat, 20)))
```

