---
title: "R Data Import"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/data-import.html)


* `read_csv()` - It creates column specification to show column's type. It is also faster than then base R read function `read.csv`.

```{r read_csv}
hotel <- read_csv("~/R/R_self study/Course_code_Coursera/hotel_bookings.csv")
```



```{r first line}
#read_csv uses the first line of the data for the column names.
read_csv("a, b, c
1, 2, 3
4, 5, 6")
```


Sometimes there are a few lines of metadata at the top of the file. You can use `skip = n` to skip the first n lines; or use `comment = "#"` to drop all lines that start with (e.g.) #.

```{r skip=n}
read_csv("The first line of metadata
         The second line of metadata
         x, y, z
         1,2,3", skip=2)
```

```{r comment}
read_csv("# A line of commment I want to skip
         x, y, z
         3, 2,1", comment = "#")
```


The data might not have column names. You can use `col_names = FALSE` to tell `read_csv()` not to treat the first row as headings, and instead label them sequentially from X1 to Xn.

```{r col_names}
read_csv("1,2,3
         13, 15, 20", col_names = F)
```

Alternatively, you can pass `col_names` a character vector which will be used as the column names:

```{r col_names2}
read_csv("1,2,3\n 15,16,20", col_names = c("x","y","z"))
```


NA in the file:

```{r NA}
read_csv("a,b,c
         1,2,3
         4,.,.", na=".")
```

