---
title: "R Tidy Data"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/tidy-data.html)


* Pivot longer

```{r pivot_longer}
table4a

pivot_longer(table4a, c(`1999`, `2000`), names_to = "year", values_to = "cases")
```

* Pivot wider

```{r pivot_wider}
table2

pivot_wider(table2, names_from = type, values_from = count)
```

Example:

```{r tribble, echo=TRUE}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes", NA, 10,
  "no", 20, 12
)

preg
```


```{r pivot_long2, echo=TRUE}
preg_tidy <- preg %>% 
  pivot_longer(c("male", "female"), names_to = "sex", values_to = "count")

preg_tidy
```



```{r value_drop_na}
#remove the (male, pregnant) row with a missing value
preg_tidy2 <- preg %>% 
  pivot_longer(c("male", "female"), names_to = "sex", values_to = "count", values_drop_na = T)

preg_tidy2
```




```{r logical, echo=TRUE}
preg_tidy3 <- preg_tidy2 %>% 
  mutate(
    female = sex == "female",
    pregnant = pregnant == "yes"
  ) %>% 
  select(pregnant, female, count)

preg_tidy3
```


#### Separating and uniting

* `separate()`

```{r seperate}
table3

table3 %>% 
  separate(rate, into = c("cases", "population"))
```


By default, `separate()` can detect non-letter/ number character and split based on that character. However, you can specify the character to split the column by using `sep` argument of `separate()`.

```{r sep}
table3 %>% separate(rate, into = c("cases", "population"), sep = "/")
```



Look careful the type of cases and population after separation, they still remain the original type of column (chr). You can use `convert=T` to change it.


```{r convert, echo=TRUE}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = T)
```


```{r sep2}
table_century <- table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

table_century
```


* `unite()`

By default, `unite()` will add underscore (_) between values from different columns.

```{r unite}
table_century %>% 
  unite(new, century, year)
```


```{r spe3}
#remove the default underscore
table_century %>% 
  unite(new, century, year, sep = "")
```


#### Missing Values

* **Explicitly**: indicated with NA.
* **Implicitly**: simply not present in the data.

```{r missing value}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```


```{r implicit}
#if the explicit NA is not important, you can hide it implicitly.
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(c(`2015`, `2016`), names_to = "year", values_to = "return", values_drop_na = T)
```

* `complete()` - important tool to make missing values explicit.

```{r complete}
stocks %>% 
  complete(year, qtr)
```


* `fill()` - can fill up the missing values by last seen value.

```{r missing value2}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
```


```{r fill}
treatment %>% 
  fill(person)
```


#### Case Study

`who` is a complicated and messy TB dataset. 

```{r who}
who
```


```{r who1}
who1 <-  who %>% 
  pivot_longer(
    cols = new_sp_m014:newrel_f65,
    names_to = "key",
    values_to = "cases",
    values_drop_na = T
  )
who1

```


```{r who2}
#to standardize the values in key column
who2 <- who1 %>% 
  mutate(key = str_replace(key, "newrel", "new_rel"))
```


```{r who3}
who3 <- who2 %>% 
  separate(key, into = c("new", "type", "sexage"), sep = "_")
who3
```


```{r who4}
#check if all new is the same value
who3 %>% 
  count(new)

#remove duplicate column
who4 <- who3 %>% select(-iso2, -iso3, -new)
who4
```


```{r who5}
who5 <- who4 %>% 
  separate(sexage, into = c("sex", "age"), sep = 1)

who5
```


```{r visualize}
who5 %>% 
  group_by(country, year, sex) %>% 
  filter(year> 1995) %>% 
  summarise(
    case = sum(cases)
  ) %>% 
  unite(country_sex, country, sex, remove = F) %>% 
  ggplot(aes(year, case, group = country_sex, color=sex))+
  geom_line()
```

