---
title: "R Data Transformation"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/transform.html)

#### Prerequisites
Set the R environment with `tidyverse` and `nycflights13`.

```{r packages}
library(tidyverse)
library(nycflights13)
```

```{r head}
head(flights)
summary(flights$dep_time)
```


#### Filter
* `filter()`

```{r filter}
(dec25 <- filter(flights, month == 12, day == 25))
```

Computers use finite precision arithmetic so remember that every number you see is an approximation. Instead of relying on ==, use `near()`:

```{r near}
1/49*49 == 1 #False

near(1/49*49, 1) #True
```


`%in%`

```{r %in%}
(nov_dec <- filter(flights, month %in% c(11,12)))
```

De Morgan's law:

* !(x & y) is the same as !x | !y 
* !(x | y) is the same as !x & !y

```{r De Morgan’s law}
#both give the same result
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```


`between()`

```{r between}
filter(flights, between(month, 7, 9)) #(month >= 7 & month <= 9)
```



`is.na()`

`filter()` only includes TRUE; it excludes both FALSE and NA values. If you want to preserve missing values, ask for them explicitly.
```{r is.na}
x <- NA
is.na(x)
```

`%%` reminder operator returns the remainder of division.

```{r %%}
c(6, 15, 25) %% 6
```
`%/%` integer division
```{r %/%}
10%/%4
```



#### Arrange
* `arrange()`

```{r arrange1}
arrange(flights, year, month, desc(day))

#Missing values are always sorted at the end.
df <- tibble(y = c(5, 8, 15, NA)) 
arrange(df, desc(y))
```


```{r arrange2}
arrange(flights, desc(is.na(dep_time)), dep_time) #use arrange() to sort all missing values to the start
```


#### Select
Below can use within select.

* `starts_with()`
* `ends_with()`
* `contains()`
* `matches()` - regular expression
* `num_range()`
* `everything()`
* `any_of()`
* `all_of()`

```{r select}
select(flights, year, month, day)
select(flights, year:day)
select(flights, -(year:day)) #Select all columns except those from year to day (inclusive)

select(flights, day, month, carrier, everything()) #change the order of column without having to specify the names of all the columns.


variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
select(flights, all_of(variables))

select(flights, starts_with("dep_"), starts_with("arr_"))
```

`rename()`
```{r rename}
rename(flights, year2013 = year)
```


* `mutate()`
* `transmute()` - only to keep the new variable

```{r mutate}
flight_small <- select(flights, month, day, distance, air_time)

mutate(flight_small, speed = distance/air_time)

transmute(flight_small, speed = distance/air_time)
```

Ranking

* `min_rank()`
* `row_number()`
* `dense_rank()`

```{r ranking}
rankme <- tibble(x = c(10, 5, 1, 5, 5))
rankme <- mutate(rankme, 
                 min_rank = min_rank(x),
                 row_number = row_number(x),
                 dense_rank = dense_rank(x))
arrange(rankme, x)
```


* `top_n()`

```{r top_n, echo=TRUE}
flight_small %>% top_n(5, air_time) %>% 
  arrange(desc(air_time))
```

* `slice()`

```{r slice}
flight_small %>% slice(1:6)
```

#### Summarise by `Group_by`

```{r summaise}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    mean = mean(dep_delay)
  )

delay_flight <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    avg_delay = mean(arr_delay)
  )

ggplot(delay_flight, mapping = aes(x= avg_delay))+
  geom_freqpoly(binwidth=10)
```


It’s always a good to include either a count `n()`, or a count of non-missing values `sum(!is.na(x))`. So that you can ensure you’re not drawing conclusions based on very small amounts of data.

```{r n()}
delay_flight <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    avg_delay = mean(arr_delay, na.rm=T),
    n = n()
  ) 

ggplot(delay_flight, aes(n, avg_delay))+
  geom_point(alpha = 1/10)
```

#### Useful summary functions

* measure of location: `median()`, `mean()`
* measure of spread: `sd()`, `IQR()`, `mad()`
* measure of rank: `min()`, `quantile(x, 0.25)`, `max()`
* measure of position: `first()`, `last()`, `nth(x, 2)`
* count: `n()`, `sum(!is.na(x))`, `n_distinct()`

#### Ungrouping

```{r ungrouping}
daily <- group_by(flights, year, month, day)

(per_daily <- daily %>% 
  summarise(
    flight = n()
  ))

per_daily %>% 
  ungroup() %>% 
  summarise(flight = n())
```

