---
title: "Factors"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/factors.html?q=factor#creating-factors)


#### Creating Factors

Let set a variable that record months:

```{r x1}
x1 <- c("Dec", "Apr", "Jan", "Mar")
```



```{r problems}
#Without using factors, it might has the problem of typo
x2 <- c("Dec", "Apr", "Jam", "Mar")

#and it does not sort in a meaningful way. 
sort(x1)
```

* To create a factor, start with creating a list of valid levels.

```{r level}
#create level
month_levels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
```


```{r factor}
#now can create a factor
y1 <- factor(x1, levels = month_levels)

y1
sort(y1)
```


Any values not in the list of level will be converted to NA.

```{r NA}
y2 <- factor(x2, levels = month_levels)

y2
```


If omit the level, it will be taken from the data in alphabetical order:

```{r omit level}
factor(x1)
```


Sometime, you may like to keep the order of level as it is appeared in the data. Then you may consider to set the level to `unique(x)` or use `fct_inorder()` after factor.

```{r unique}
#unique(x)
f1 <- factor(x1, levels = unique(x1))
f1

#fct_inorder()
f2 <- x1 %>% factor() %>% fct_inorder()
f2
```


If you ever need to access the valid level of factor, you can use `levels()` directly:

```{r access level}
levels(f1)
```

#### Example of using factor

We can use the sample data from General Social Survey to study factor.


```{r sample}
gss_cat
```



When factor is stored in tibble, it is hard to see the level, you need to use `count()` or bar chart.

```{r count, echo=TRUE}
gss_cat %>% 
  count(race)
```


```{r geom_bar, echo=TRUE}
ggplot(gss_cat, aes(race))+
  geom_bar()
```


```{r drop, echo=TRUE}
#by default, ggplot will drop level that doesn't have any values, you can force it to display
ggplot(gss_cat, aes(race))+
  geom_bar()+
  scale_x_discrete(drop = F)
```



#### Modifying Factor Order

By using `fct_reorder()` which take 3 arguments to reorder the factor order:


```{r fct_reorder}
#before reorder the factor, it is hard to see any pattern
relig_tv <- gss_cat %>% 
  group_by(relig) %>% 
  summarise(
    age = mean(age, na.rm=T),
    tv_hour = mean(tvhours, na.rm=T),
    n = n()
  ) 

ggplot(relig_tv, aes(tv_hour, relig))+
  geom_point()

#use fct_reorder to reorder the factor and the pattern can be seen easily now
ggplot(relig_tv, aes(tv_hour, fct_reorder(relig, tv_hour)))+
  geom_point()
```

Using `fct_infreq()` to order levels in increasing frequency, you may consider to couple it with `fct_rev()`.

```{r fct_infreq}
gss_cat %>% 
  mutate(marital = fct_infreq(marital) %>% fct_rev()) %>% 
  ggplot(aes(marital))+
  geom_bar()
```


You may use `fct_relevel()` to take any number of levels that you want to move to the front of the line.


```{r fct_relevel}
#to move "Don't know" to the first
ggplot(relig_tv, aes(tv_hour, fct_relevel(relig, "Don't know")))+
  geom_point()
```



#### Modifying Factor Level


* Using `fct_recode()` to rename the level in a factor.


```{r fct_recode}
#simply plot the rincome and it is hard to read the x-axis
ggplot(gss_cat, aes(rincome))+
  geom_bar()

#use fct_recode to rename the level
gss_cat %>% 
  filter(rincome != c("Not applicable")) %>% 
  mutate(rincome = fct_recode(rincome, "Lower than $1000" = "Lt $1000")) %>% 
  mutate(rincome_na = rincome %in% c("Refused", "Don't know", "No answer")) %>% 
  ggplot(aes(rincome, fill = rincome_na))+
  geom_bar()+
  coord_flip()+
  theme(legend.position = "none")+
  scale_x_discrete("Respondent Income")+
  scale_y_continuous("Number of Respondent", labels = scales::comma)
```


* Using `fct_collapse` as a variation of `fct_recode` to collapse many levels.

```{r fct_collapse}
gss_cat %>% 
  count(partyid)

gss_cat %>% 
  mutate(partyid = fct_collapse(partyid,
   other = c("No answer", "Don't know", "Other party"),
   rep = c("Strong republican", "Not str republican"),
   ind = c("Ind,near rep", "Independent", "Ind,near dem"),
   dem = c("Not str democrat", "Strong democrat")
  )) %>% 
  count(partyid)
```

* Using `fct_lump` to to lump together all the small groups and specify `n` parameter on how many groups (excluding other) we want to keep.

```{r}
gss_cat %>% 
  mutate(partyid = fct_lump(partyid, n = 5)) %>% 
  count(partyid, sort = T)
```


