---
title: "R Dates and Times"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/dates-and-times.html)


#### Lubridate

`lubridate` is not part of core `tidyverse`, we load it only when we want to deal with date/ time.

```{r lubridate}
library(lubridate)
library(tidyverse)
library(nycflights13)
```


#### Creating date/times
* date 
* time
* date-time: in tibble (dttm), elsewhere in R is called `POSIXct`.

```{r today}
#to get current date/ date-time
today()

now()
```


#### Create date/time from string

```{r create date}
#create date from string
ymd("2017-01-31")
#> [1] "2017-01-31"

mdy("January 31st, 2017")
#> [1] "2017-01-31"

dmy("31-Jan-2017")
#> [1] "2017-01-31"

#this function also takes unquoted numbers
ymd(20170131)
#> [1] "2017-01-31"
```


```{r create date-time}
#create date-time
ymd_hms("2017-01-31 20:11:59")
#> [1] "2017-01-31 20:11:59 UTC"

mdy_hm("01/31/2017 08:01")
#> [1] "2017-01-31 08:01:00 UTC"

#by providing time zone
ymd(20170131, tz = "UTC")
#> [1] "2017-01-31 UTC"
```


#### Create date/time from individual components

Individual components of date and time are spread across multiple columns in flights.

```{r flights}
flights %>% 
  select(year, month, day, hour, minute)
```


By using `make_date` or `make_datetime` to create a date/time from this kind of input.


```{r make_datetime}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate("depature" = make_datetime(year, month, day, hour, minute))
```


#### Example

```{r example}
#use modulus arithmetic to pull out the hour and minute components
make_datetime_100 <- function(year, month, day, time){
  make_datetime(year, month, day, time%/%100, time%%100)
}

flight_dt <- flights %>%
  filter(!is.na(arr_time), !is.na(dep_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flight_dt

#visualize depart time across a year
ggplot(flight_dt, aes(dep_time))+
  geom_freqpoly(binwidth=86400) #86400s = 1 day

#visualize within a single day
flight_dt %>% 
  filter(dep_time < ymd(20130102)) %>% 
  ggplot(aes(dep_time))+
  geom_freqpoly(binwidth=600) #600s = 10 min
```


#### Create date/time from other type

`as_date()` and `as_datetime()`

```{r as_date}
as_date(now())

as_datetime(today())
```


#### Date-time component

* Getting component:
    * `year()`
    * `month()`
    * `mday()` (day of the month)
    * `yday()` (day of the year)
    * `wday()` (day of the week)
    * `hour()`, `minute()`, `second()`
    
```{r getting component}
datetime <- now()

year(datetime)
month(datetime)
day(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
hour(datetime)
minute(datetime)
second(datetime)
```


Set `label = TRUE` to return the abbreviated name of the month or day of the week. Set `abbr = FALSE` to return the full name.

```{r label}
month(datetime, label = T)
```

```{r abbr}
wday(datetime, label = T, abbr = F)
```


We can see more flights on the week than weekend:

```{r wday}
flight_dt %>% 
  mutate(wday = wday(dep_time, label = T)) %>% 
  ggplot(aes(wday))+
  geom_bar()
```



You can create a new date-time with `update()` which also allows you to set multiple values at once.

```{r update}
update(datetime, year = 2022, month = 2, mday = 2, hour = 2)
```

#### Time spans

* `durations`, which represent an exact number of seconds.
* `periods`, which represent human units like weeks and months.
* `intervals`, which represent a starting and ending point.


When subtract two dates, you get a difftime object. However, this difftime object records time spans in seconds, minutes, hours, days, or weeks, which cause ambiguous. So, you can use lubridate function of `duration` which always record in second only.

```{r difftime}
h_age <- today() - ymd(19900521)
h_age
```

```{r duration}
as.duration(h_age)
```

Durations come with a bunch of convenient constructors:

```{r}
dseconds(20)
dminutes(10)
ddays(3)
dhours(c(1,5))
dweeks(0:3)
dyears(1)
```

