---
title: "r4ds_rmarkdown_data visualisation_1"
author: "KAR RUI, TAN"
date: "10/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Reference
[Link](https://r4ds.had.co.nz/data-visualisation.html#data-visualisation)

#### Prerequisites
Set the R environment with `tidyverse`.

```{r tidyverse}
library(tidyverse)

```

#### Creating a `ggplot()`
Using `mpg` data to shows a negative relationship between engine size (displ) and fuel efficiency (hwy).

```{r ggplot}

head(mpg)

ggplot(mpg, aes(displ, hwy))+
  geom_point()

```

#### Geometric objects
Useless to use scatter plot on two categorical variables (drv & class). Replace `geom_point` with `geom_count` or `geom_tile`.

* `geom_count`
```{r geom_count}

ggplot(mpg, aes(class, drv))+
  geom_count()

```

* `geom_tile`
```{r geom_tile}

count(mpg, class, drv) %>% 
  ggplot(aes(class, drv))+
  geom_tile(aes(fill = n))

```

* `geom_smooth`

```{r geom_smooth}
ggplot(mpg, aes(displ, hwy))+
  geom_smooth(aes(linetype=drv, color=drv))+
  geom_point(aes(color=drv))


ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(data = filter(mpg, class == "subcompact"), se=F)

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```


* `geom_bar`
Every geom has a default stat and every stat has a default geom. You can generally use geoms and stats interchangeably.

```{r geom_bar}
ggplot(diamonds, aes(cut))+
  geom_bar()

#another way of using stat to plot the same geom
ggplot(diamonds)+
  stat_count(aes(x=cut))
```


* `geom_jitter`

```{r geom_jitter}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter(width = 1, height = 2)
```


* `geom_boxplot`

```{r geom_boxplot}
ggplot(mpg, aes(drv, hwy))+
  geom_boxplot()
```


* `geom_histogram`

```{r geom_histogram}
ggplot(diamonds, aes(carat))+
  geom_histogram(binwidth = 0.5)
```


* `geom_freqpoly`

```{r geom_freqpoly}
ggplot(diamonds, aes(carat, color=cut))+
  geom_freqpoly(binwidth = 0.1)
```


#### Aesthetic mapping
* size
* shape - max 6 variables
* color
* alpha - transparency
* stroke - changes the size of the border for shapes (21-25)

```{r aes}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class, shape=class))
```

```{r stroke}
ggplot(mpg, aes(displ, hwy))+
  geom_point(color="white", shape=24, fill="pink", stroke=1, size=3)
```

Boolean condition: 
```{r boolean }
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color= displ<5 | hwy<=20))
```


#### Facet
* `facet_wrap`

```{r facet_wrap}
ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  facet_wrap(~class,nrow = 4)

ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  facet_wrap(~class ~drv)
```

* `facet_grid`
```{r facet_grid}
ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  facet_grid(class~drv)

ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  facet_grid(.~drv)

ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  facet_grid(drv~.)
```


#### Stastical Transformation
Three reasons you might need to use a stat explicitly:

* 1) Override the default stat. Eg: change 'count' (default) to 'identity' in `geom_bar`.
```{r stat(identity)}
demo <- tribble(
   ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

ggplot(demo)+
  geom_bar(aes(cut, freq), stat = "identity")
```

* 2) Override the default mapping from transformed variables to aesthetics. Eg: change 'count' to 'prop'.
```{r stat(prop)}
ggplot(diamonds)+
  geom_bar(aes(cut, stat(prop), group=1))
```

* 3) Draw bigger attention to the statistical transformation in your code.
```{r stat_summary}
ggplot(diamonds)+
  stat_summary(
    aes(cut, depth),
    fun.max = max,
    fun.min = min,
    fun = median
  )

#another way of using geom to plot the same plot
ggplot(diamonds)+
  geom_pointrange(
    aes(cut, depth),
    stat = "summary",
    fun.max = max,
    fun.min = min,
    fun = median
  )
```


#### Position Adjustment
* Position: "stack" (default in `geom_bar`)
```{r stack position}
ggplot(diamonds, aes(cut))+
  geom_bar(aes(fill=clarity))
```

* Position: "identity", will place each object exactly where it falls in the context of the graph.
```{r identity position}
ggplot(diamonds, aes(cut, color=clarity))+
  geom_bar(fill = NA, position = "identity")
```

* Position: "fill", makes each set of stacked bars the same height.
```{r fill position}
ggplot(diamonds, aes(cut, fill=clarity))+
  geom_bar(position = "fill")
```

* Position: "dodge", places overlapping objects directly beside one another.
```{r dodge position}
ggplot(diamonds, aes(cut, fill=clarity))+
  geom_bar(position = "dodge")
```

* Position: "jitter", adds a small amount of random noise to each point. To solve **overplotting** problem for points.
```{r jitter position}
#see geom_jitter as well
ggplot(mpg, aes(displ, hwy))+
  geom_point(position = "jitter")
```


#### Coordinate System
* `Coord_flip`
```{r coord_flip}
ggplot(mpg, aes(drv, hwy))+
  geom_boxplot()+
  coord_flip()
```

* `Coord_quickmap`
```{r coord_quickmap}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group=group))+
  geom_polygon(fill="white", color="black")+
  coord_quickmap()
```

* `Coord_polar`
```{r coord_polar}
ggplot(diamonds)+
  geom_bar(aes(cut, fill=cut), show.legend = F)+
  labs(x=NULL, y=NULL)+
  coord_polar()

#theta = "y" maps y to the angle of each section
ggplot(mpg, aes(x=factor(1), fill=drv))+
  geom_bar()+
  coord_polar(theta = "y")
```

* `Coord_fixed` - ensures that the line produced by geom_abline() is at a 45-degree angle.
```{r coord_fixed}
ggplot(mpg, aes(cty, hwy))+
  geom_point()+
  geom_abline()+
  coord_fixed()
```

* `Coord_cartesian` - zoom to small values of the x (xlim) or y-axis (ylim).
```{r coord_cartesian}
ggplot(diamonds, aes(x=y))+
  geom_histogram(binwidth = 0.5)+
  coord_cartesian(ylim = c(0,50))
```


#### Reference
[Link](https://r4ds.had.co.nz/graphics-for-communication.html)



#### Label

```{r label}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color = class))+
  geom_smooth(se=F)+
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weigh",
    caption = "Data from fueleconomy.gov",
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type"
  )
```


#### Annotation

* `geom_text` - behave like `geom_point` but allow labeling text on the plot.


```{r geom_text}
best_in_class <- mpg %>% 
  group_by(class) %>% 
  filter(row_number(desc(hwy))==1)

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color = class))+
  geom_text(aes(label = model), data = best_in_class)
```



* `geom_label` - draws a rectangle behind the text. You can use `nudge_y` parameter to move the labels slightly above the corresponding points.


```{r geom_label}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color = class))+
  geom_label(aes(label = model), data = best_in_class, nudge_y = 2, alpha = 0.3)
```



#### Legend

```{r legend}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  theme(legend.position = "none") #or "top", "bottom", "left", "right"(default)
```



#### Scales

Behind the scene, ggplot2 automatically adds default scales:
* `scale_x_continuous`
* `scale_y_continuous`
* `scale_color_discrete`

The formula is scale_, followed by name of aesthetics, then _, then the name of the scale (continuous, discrete, datetime, or date).

There are two primary arguments: `break` and `label`.

```{r scales}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  scale_x_continuous(labels = NULL)+
  scale_y_continuous(breaks = seq(15,40, by=5))
```



Manually set the color.

```{r scale_color_manual}
ggplot(mpg, aes(displ, hwy, color = drv))+
  geom_point()+
  scale_color_manual(values = c(`4`= "pink", f = "blue", r = "yellow"))
```


For continuous color, you can use:

* `scale_color_gradient`
* `scale_fill_gradient`

```{r scale_fill_gradient}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

ggplot(df, aes(x,y))+
  geom_hex()+
  coord_fixed()+
  scale_fill_gradient(low = "white", high = "red")
```



#### Zooming

* `coord_cartesian`

```{r coord_cartesian2}
ggplot(mpg, mapping = aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 7), ylim = c(10, 30))
```


```{r fixed scales for two plot comparison}
#this example show it's hard to compare two plot with different x-axis, y-axis, and color aesthetic with different range.
suv <- mpg %>% filter(class == "suv")
compact <- mpg %>% filter(class == "compact")

ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point()

ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point()

#other than faceting to solve the problem, we can use the below solution if we want to have 2 plots for comparison.
x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_colour_discrete(limits = unique(mpg$drv))

ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale

ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale
```

