---
title: "R Tibbles"
output: html_notebook
---

#### Reference
[Link](https://r4ds.had.co.nz/tibbles.html)


#### Creating Tibbles

* `as_tibble`

```{r as_tibble}
as_tibble(iris)
```

* `tibble`
```{r tibble}
tibble(
  x = 1:5,
  y = 2,
  z = x^2+y
)
```

* `tribble` - transposed tibble. The column headings are defined by formulas (start with ~), and entries are separated by commas.

```{r tribble}
#add a commented line to know where the header is
tribble(
  ~x, ~y, ~z,
 #--| --| --|
 "a", 2, 32,
 "b", 1, 15
)
```



It’s okay for a tibble to have column names that are not valid R variable names, aka non-syntactic names. For example, they might not start with a letter, or they might contain unusual characters like a space. You need to surround them with backticks, `:

```{r backtick}
tb <- tibble(
  `:)` = "smile",
  `200` = "number"
)

tb
```


#### Subsetting

`[[` can extract by name or position; `$` only extracts by name but is a little less typing.

```{r subsetting}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x

df[["x"]]

# Extract by position
df[[1]]
```

